---
title: "Sample github_document"
author: "Zhenguo Zhang"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  github_document:
    toc: yes
    toc_depth: 3
  html_document:
    toc: yes
    toc_depth: '3'
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

In this demo, we shows how to create a github-flavored markdown
document, which can be viewed as html at github directly.

## Requirements

* latest (*rmarkdown*)[https://rmarkdown.rstudio.com/] package: the *github_document* format is only available in recent *rmarkdown* package.

## The document

Here we go. Nothing is special when creating github_document, just
write the document as writing html_document, and *rmarkdown::render()*
will take care of the rest.

Here we create a plot using the builtin data *cars*:

```{r demo-plot, fig.cap= "Demo plot"}
with(cars, plot(speed ~ dist, pch=20, col="red"))
```

## Compile to github_document

When a Rmarkdown document is complete, use *rmarkdown::render()* or
the "Knit" button to generate the output markdown document. This
yields a markdown file with *.md* extension as well as a folder containing
elements for the document such as generated figures.

Push the markdown as well as associated folder to github and
then you can view this markdown file at github.

## Options for github_document

Here are useful options can be put in the YAML header to tune
the outlook of the github document.

Option | Values | Explanation
--- | --- | :---
html_preview | true, false | if false, suppress html preview
toc | true, false | if false, no table of contents
toc_depth | integer | default 3, the depth of sections in TOC
fig_width | float | default 7, the width of output figures
fig_height | float | default 5, the height of output figures
dev | jpeg, png, etc | default png, device for figure output.
includes | in_header, before_body, after_body | include other markdown files in header, before or after document body
md_extensions | +pandoc_opt, -pandoc_opt | enable/disable pandoc extensions
pandoc_args | ["--opt","value"] | use pandoc options directly

To share options among multiple markdown documents in a directory,
one can put these options into file *_output.yaml*.

Have fun!

## References

1. https://rmarkdown.rstudio.com/github_document_format.html

